---
title: "Cardiomyopathy WGCNA"
author: "Brandon Teo"
date: "2026-03-06"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("intersect", "base")
conflicts_prefer(dplyr::slice)
```

### Background

Dilated cardiomyopathy (DCM) is the most common non-ischaemic cardiomyopathy and a leading cause of heart failure and transplantation. Disease progression is driven in part by maladaptive extracellular matrix (ECM) remodelling and cardiac fibrosis, yet the co-ordinated transcriptional programmes underlying these processes remain poorly characterised.
Standard differential expression analysis identifies individual genes that differ between conditions but treats each gene independently, offering limited insight into the higher-order regulatory networks that govern complex phenotypes like fibrosis. Weighted Gene Co-expression Network Analysis (WGCNA) addresses this limitation by grouping genes into modules of co-expressed transcripts and summarising each module with a single eigengene. Module eigengenes can then be correlated with clinical traits, such as disease status, to identify gene programmes collectively associated with DCM. Hub genes within trait-associated modules, defined by high intramodular connectivity, represent candidate regulators of disease-relevant processes.
Thus, we apply WGCNA to bulk RNA-seq data from two independent studies of human left ventricular myocardium (GSE116250 and GSE141910), focusing exclusively on DCM patients and non-failing controls. We construct a co-expression network, identify trait-associated modules, and characterise them through pathway enrichment analysis with particular attention to ECM organisation and TGF-β signalling. This network-based approach complements conventional differential expression by providing a systems-level view of the transcriptomic landscape of DCM-associated cardiac fibrosis.

Research question: What co-expressed gene modules are robustly associated with dilated cardiomyopathy across independent bulk RNA-seq cohorts, and do module-level transcriptional programs implicate fibrotic, inflammatory, or metabolic dysregulation as primary drivers of disease?

---

## Step 1: Load Libraries

```{r install_bioc_packages, message=FALSE, warning=FALSE}
# Install Bioconductor packages if not already present
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

bioc_pkgs <- c("WGCNA", "limma", "GEOquery", "AnnotationDbi", "org.Hs.eg.db", "sva")
for (pkg in bioc_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE))
    BiocManager::install(pkg)
}
```

```{r load_libraries, message=FALSE, warning=FALSE}
library(WGCNA)
library(limma)
library(GEOquery)
library(tidyverse)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(sva)
library(gridExtra)

# Allow multi-threading for WGCNA
allowWGCNAThreads()
```

## Step 2: Load Metadata (Series Matrices)

```{r load_metadata, message=FALSE, warning=FALSE}
getwd()
# Load GEO series matrix files, these contain sample-level metadata only
gse116250_geo <- getGEO(filename = "GSE116250_series_matrix.txt",    GSEMatrix = TRUE)
gse141910_geo <- getGEO(filename = "GSE141910_series_matrix.txt.gz", GSEMatrix = TRUE)

# Extract phenotypic data tables (rows = samples, columns = metadata fields)
pheno_gse116250 <- pData(phenoData(gse116250_geo))
pheno_gse141910 <- pData(phenoData(gse141910_geo))

cat("GSE116250 total samples:", nrow(pheno_gse116250), "\n")
cat("GSE141910 total samples:", nrow(pheno_gse141910), "\n")
```

## Step 2b: Quality Check — Metadata Condition Labels

Before filtering, inspect the **exact** strings in the condition columns. This catches
capitalisation differences, trailing whitespace, or unexpected categories that would
silently drop samples if the filter string doesn't match.

```{r qc_metadata_labels}
# Print the exact unique values, what you see here is what you must type in the filter
cat("=== GSE116250 — unique values in 'disease:ch1' ===\n")
unique_116250 <- unique(trimws(pheno_gse116250[["disease:ch1"]]))
print(unique_116250)

# Detect any leading/trailing whitespace (invisible but breaks string matching)
has_whitespace_116250 <- any(pheno_gse116250[["disease:ch1"]] != trimws(pheno_gse116250[["disease:ch1"]]))
cat("Whitespace detected in GSE116250 disease column:", has_whitespace_116250, "\n\n")

cat("=== GSE141910 — unique values in 'etiology:ch1' ===\n")
unique_141910 <- unique(trimws(pheno_gse141910[["etiology:ch1"]]))
print(unique_141910)

has_whitespace_141910 <- any(pheno_gse141910[["etiology:ch1"]] != trimws(pheno_gse141910[["etiology:ch1"]]))
cat("Whitespace detected in GSE141910 etiology column:", has_whitespace_141910, "\n")

# Print all available column names to identify which fields store age and sex.
# GEO deposits clinical variables as "characteristic:ch1" columns — the exact
# spelling varies by submitter (e.g. "age:ch1", "Age:ch1", "sex:ch1", "gender:ch1").
# Review the output here; the extraction code in Step 3 auto-detects by pattern.
cat("\n=== GSE116250 — all phenotype column names ===\n")
print(colnames(pheno_gse116250))

cat("\n=== GSE141910 — all phenotype column names ===\n")
print(colnames(pheno_gse141910))
```

## Step 3: Filter Metadata to DCM and Non-Failing Samples Only

```{r filter_metadata}
# GSE116250
# Condition is stored in the 'disease:ch1' column
# Values: "dilated cardiomyopathy", "ischemic cardiomyopathy", "non-failing"
cat("GSE116250 disease groups:\n")
print(table(pheno_gse116250[["disease:ch1"]]))

pheno_gse116250 <- pheno_gse116250 %>%
  filter(`disease:ch1` %in% c("dilated cardiomyopathy", "non-failing")) %>%
  mutate(condition = ifelse(`disease:ch1` == "non-failing", "NF", "DCM"))

cat("\nGSE116250 after filter:", nrow(pheno_gse116250), "samples\n")
print(table(pheno_gse116250$condition))

# Extract age and sex from GSE116250
# Auto-detect the column names; GEO naming is submitter-defined
age_col_116250 <- grep("^age",         colnames(pheno_gse116250), ignore.case = TRUE, value = TRUE)[1]
sex_col_116250 <- grep("^(sex|gender)", colnames(pheno_gse116250), ignore.case = TRUE, value = TRUE)[1]
cat("GSE116250 age column detected:", age_col_116250, "\n")
cat("GSE116250 sex column detected:", sex_col_116250, "\n")

# Age: coerce to numeric (suppresses warning if any values are non-numeric / "NA")
pheno_gse116250$age <- suppressWarnings(as.numeric(trimws(pheno_gse116250[[age_col_116250]])))

# Sex: standardise to Male = 1, Female = 0, anything else = NA
sex_raw_116250 <- trimws(tolower(pheno_gse116250[[sex_col_116250]]))
pheno_gse116250$sex <- dplyr::case_when(
  sex_raw_116250 == "male"   ~ 1L,
  sex_raw_116250 == "female" ~ 0L,
  TRUE                       ~ NA_integer_
)
cat("GSE116250 age summary: mean =", round(mean(pheno_gse116250$age, na.rm = TRUE), 1),
    " NAs =", sum(is.na(pheno_gse116250$age)), "\n")
cat("GSE116250 sex (1=M, 0=F, NA=unknown):", "\n")
print(table(pheno_gse116250$sex, useNA = "ifany"))

# GSE141910
# Condition is stored in the 'etiology:ch1' column
# Values: "Dilated cardiomyopathy (DCM)", "Non-Failing Donor",
#         "Hypertrophic cardiomyopathy (HCM)", "Peripartum cardiomyopathy (PPCM)"
cat("\nGSE141910 etiology groups:\n")
print(table(pheno_gse141910[["etiology:ch1"]]))

pheno_gse141910 <- pheno_gse141910 %>%
  filter(`etiology:ch1` %in% c("Dilated cardiomyopathy (DCM)", "Non-Failing Donor")) %>%
  mutate(condition = ifelse(`etiology:ch1` == "Non-Failing Donor", "NF", "DCM"))

cat("\nGSE141910 after filter:", nrow(pheno_gse141910), "samples\n")
print(table(pheno_gse141910$condition))

#Extract age and sex from GSE141910
age_col_141910 <- grep("^age",         colnames(pheno_gse141910), ignore.case = TRUE, value = TRUE)[1]
sex_col_141910 <- grep("^(sex|gender)", colnames(pheno_gse141910), ignore.case = TRUE, value = TRUE)[1]
cat("GSE141910 age column detected:", age_col_141910, "\n")
cat("GSE141910 sex column detected:", sex_col_141910, "\n")

pheno_gse141910$age <- suppressWarnings(as.numeric(trimws(pheno_gse141910[[age_col_141910]])))

sex_raw_141910 <- trimws(tolower(pheno_gse141910[[sex_col_141910]]))
pheno_gse141910$sex <- dplyr::case_when(
  sex_raw_141910 == "male"   ~ 1L,
  sex_raw_141910 == "female" ~ 0L,
  TRUE                       ~ NA_integer_
)
cat("GSE141910 age summary: mean =", round(mean(pheno_gse141910$age, na.rm = TRUE), 1),
    " NAs =", sum(is.na(pheno_gse141910$age)), "\n")
cat("GSE141910 sex (1=M, 0=F, NA=unknown):", "\n")
print(table(pheno_gse141910$sex, useNA = "ifany"))
```

## Step 4: Load Count Matrices

### 4a. GSE116250 — Pre-assembled raw integer counts

```{r load_gse116250_counts}
# This file is a tab-separated matrix:
#   rows    = genes (Entrez IDs)
#   columns = samples (GSM accession IDs)
#   values  = raw read counts from STAR/featureCounts
counts_gse116250 <- read.table(
  "GSE116250_raw_counts_GRCh38.p13_NCBI.tsv",
  header      = TRUE,
  sep         = "\t",
  row.names   = 1,
  check.names = FALSE
)

cat("GSE116250 count matrix dimensions:", dim(counts_gse116250), "\n")
cat("Gene ID format (Entrez):", head(rownames(counts_gse116250), 3), "\n")
cat("Sample ID format (GSM):", head(colnames(counts_gse116250), 3), "\n")
cat("Values are integers:", !any(as.matrix(counts_gse116250) != round(as.matrix(counts_gse116250)), na.rm = TRUE), "\n")
```

### 4b. GSE141910 — Assemble from individual per-sample files

**Note:** The GEO-deposited files for GSE141910 contain Voom-transformed log2 values
(confirmed by GEO metadata: *"Raw gene counts were transformed using SVA and the
LIMMA-Voom method"*). Raw STAR counts were not deposited. These pre-normalised values
will be used directly for WGCNA after filtering; 

```{r assemble_gse141910_counts, message=FALSE}
folder_path <- "GSE141910_counts"
file_list   <- list.files(path = folder_path, pattern = "\\.gz$", full.names = TRUE)
cat("Number of sample files found:", length(file_list), "\n")

# Function to read one sample file and name its count column with the GSM accession ID
read_and_name <- function(file) {
  temp_data   <- read.csv(gzfile(file), header = TRUE)
  # Extract the GSM accession ID from the filename (everything before the first underscore)
  sample_name <- gsub("(_.*|\\.csv\\.gz|\\.gz)", "", basename(file))
  colnames(temp_data) <- c("Gene", sample_name)
  return(temp_data)
}

# Read all files and merge into a single matrix by the Gene column
# Note all_samples_list takes a while to run as there are many folders to merge, may take 10minutes
# all = FALSE: inner join — keeps only genes present in every sample, avoiding NAs
all_samples_list <- lapply(file_list, read_and_name)
counts_gse141910 <- Reduce(
  function(x, y) merge(x, y, by = "Gene", all = FALSE),
  all_samples_list
)

# Promote gene IDs to row names
rownames(counts_gse141910) <- counts_gse141910$Gene
counts_gse141910 <- counts_gse141910[, -1]

cat("GSE141910 count matrix dimensions:", dim(counts_gse141910), "\n")
cat("Gene ID format (ENSEMBL):", head(rownames(counts_gse141910), 3), "\n")
cat("Sample ID format (GSM):", head(colnames(counts_gse141910), 3), "\n")
```

## Step 4c: Quality Check — Count Matrices

```{r qc_count_matrices}
# GSE116250
cat("=== GSE116250 count matrix ===\n")
cat("Dimensions (genes x samples):", dim(counts_gse116250), "\n")
cat("Values are integers:", !any(as.matrix(counts_gse116250) != round(as.matrix(counts_gse116250)), na.rm = TRUE), "\n")
cat("Missing values (NA):", sum(is.na(counts_gse116250)), "\n")
cat("Value range: min =", min(as.matrix(counts_gse116250), na.rm = TRUE),
    "  max =", max(as.matrix(counts_gse116250), na.rm = TRUE), "\n")

# Check that every sample in the filtered metadata exists in the count matrix
missing_116250 <- base::setdiff(rownames(pheno_gse116250), colnames(counts_gse116250))
cat("Metadata samples missing from count matrix:", length(missing_116250), "\n")
if (length(missing_116250) > 0) print(missing_116250)

# Preview top-left corner
cat("\nFirst 5 genes x 4 samples:\n")
print(counts_gse116250[1:5, 1:4])

# GSE141910
cat("\n=== GSE141910 count matrix ===\n")
cat("Dimensions (genes x samples):", dim(counts_gse141910), "\n")
cat("Values are non-integer (Voom log2):", any(as.matrix(counts_gse141910) != round(as.matrix(counts_gse141910)), na.rm = TRUE), "\n")
cat("Missing values (NA):", sum(is.na(counts_gse141910)), "\n")
cat("Value range: min =", round(min(as.matrix(counts_gse141910), na.rm = TRUE), 2),
    "  max =", round(max(as.matrix(counts_gse141910), na.rm = TRUE), 2), "\n")

missing_141910 <- base::setdiff(rownames(pheno_gse141910), colnames(counts_gse141910))
cat("Metadata samples missing from count matrix:", length(missing_141910), "\n")
if (length(missing_141910) > 0) print(missing_141910)

cat("\nFirst 5 genes x 4 samples:\n")
print(round(counts_gse141910[1:5, 1:4], 3))
```

## Step 5: Quality Check — Sample Name Matching

Run this **before** alignment. If any mismatches are found here, fix them before
proceeding — the alignment step will crash without useful output if names don't match.

```{r qc_sample_matching}
# GSE116250

cat("=== GSE116250 sample name matching ===\n\n")

# Side-by-side: first 6 metadata IDs vs count matrix column names
cat("First 6 metadata rownames vs count matrix column names:\n")
print(data.frame(
  metadata_ID = head(rownames(pheno_gse116250)),
  countmat_ID = head(colnames(counts_gse116250))
))

# Any samples in filtered metadata absent from count matrix?
only_in_meta_116250 <- base::setdiff(rownames(pheno_gse116250), colnames(counts_gse116250))
cat("\nIn metadata but NOT in count matrix:", length(only_in_meta_116250), "\n")
if (length(only_in_meta_116250) > 0) print(only_in_meta_116250)

# Any extra columns in count matrix with no metadata entry?
only_in_counts_116250 <- base::setdiff(colnames(counts_gse116250), rownames(pheno_gse116250))
cat("In count matrix but NOT in metadata:", length(only_in_counts_116250), "\n")
if (length(only_in_counts_116250) > 0) print(only_in_counts_116250)

cat("\nExpected condition breakdown (from metadata):\n")
print(table(pheno_gse116250$condition))

# GSE141910

cat("\n=== GSE141910 sample name matching ===\n\n")

cat("First 6 metadata rownames vs count matrix column names:\n")
print(data.frame(
  metadata_ID = head(rownames(pheno_gse141910)),
  countmat_ID = head(colnames(counts_gse141910))
))

only_in_meta_141910 <- base::setdiff(rownames(pheno_gse141910), colnames(counts_gse141910))
cat("\nIn metadata but NOT in count matrix:", length(only_in_meta_141910), "\n")
if (length(only_in_meta_141910) > 0) print(only_in_meta_141910)

only_in_counts_141910 <- base::setdiff(colnames(counts_gse141910), rownames(pheno_gse141910))
cat("In count matrix but NOT in metadata:", length(only_in_counts_141910), "\n")
if (length(only_in_counts_141910) > 0) print(only_in_counts_141910)

cat("\nExpected condition breakdown (from metadata):\n")
print(table(pheno_gse141910$condition))
```

## Step 6: Align Count Matrices with Filtered Metadata

```{r align_gse116250}
# Subset count matrix columns to the filtered samples, reordered to match metadata rows
counts_gse116250 <- counts_gse116250[, rownames(pheno_gse116250)]

# Hard stop: every column must exactly match the corresponding metadata row
stopifnot(all(colnames(counts_gse116250) == rownames(pheno_gse116250)))
cat("GSE116250 alignment check passed:",
    ncol(counts_gse116250), "samples,",
    nrow(counts_gse116250), "genes.\n")
```

```{r align_gse141910}
counts_gse141910 <- counts_gse141910[, rownames(pheno_gse141910)]

stopifnot(all(colnames(counts_gse141910) == rownames(pheno_gse141910)))
cat("GSE141910 alignment check passed:",
    ncol(counts_gse141910), "samples,",
    nrow(counts_gse141910), "genes.\n")
```

## Step 7: Normalise GSE116250 to log2-CPM

GSE116250 raw integer counts are filtered then converted to log2-CPM, placing them on
the same scale as the GSE141910 Voom log2-CPM values for merging.

```{r normalise_gse116250, message=FALSE, warning=FALSE}
# Filter: remove genes with counts < 15 in more than 75% of samples (WGCNA RNA-seq FAQ)
# Equivalently: keep genes where count >= 15 in at least 25% of samples
keep <- rowSums(counts_gse116250 >= 15) >= ceiling(ncol(counts_gse116250) * 0.25)
filtered116250 <- as.matrix(counts_gse116250[keep, ])
cat("GSE116250 genes after low-count filter:", nrow(filtered116250), "\n")

# log2-CPM normalisation
# lib.size: total mapped reads per sample — used to make samples comparable
# regardless of sequencing depth. +0.5 / +1 are pseudocounts to avoid log(0).
lib.size <- colSums(filtered116250)
y <- t(log2(t(filtered116250 + 0.5) / (lib.size + 1) * 1e6))

# Boxplot: raw counts vs log2-CPM — distributions should be roughly even across samples
par(mfrow = c(1, 2))
boxplot(log2(filtered116250 + 1),
        las = 2, main = "Raw counts (log2)", ylab = "log2(count+1)")
boxplot(y,
        las = 2, main = "log2-CPM normalised", ylab = "log2-CPM")

expr_gse116250 <- y
cat("GSE116250 expression matrix dimensions:", dim(expr_gse116250), "\n")
cat("Value range:", round(min(expr_gse116250), 2), "—", round(max(expr_gse116250), 2), "\n")
```

## Step 7b: Filter GSE141910 Pre-normalised Values

GSE141910 Voom values are on a log2-CPM scale. The WGCNA FAQ threshold of 15 raw counts
does not apply here. The standard limma/edgeR convention for log2-CPM data is to retain
genes with log2-CPM > 1 (i.e. CPM > 2) in at least 25% of samples — equivalent to
removing genes absent in more than 75% of samples.

```{r filter_gse141910}
keep141910   <- rowSums(counts_gse141910 >= 1) >= ceiling(ncol(counts_gse141910) * 0.25)
expr_gse141910 <- as.matrix(counts_gse141910[keep141910, ])

cat("GSE141910 genes after expression filter:", nrow(expr_gse141910), "\n")
cat("Value range:", round(min(expr_gse141910), 2), "—", round(max(expr_gse141910), 2), "\n")
```

## Step 8: Harmonise Gene IDs to HGNC Gene Symbols

GSE116250 uses Entrez IDs; GSE141910 uses Ensembl IDs. Both are mapped to HGNC symbols
so the two matrices share a common identifier for merging.

```{r harmonise_ids_gse116250}
# GSE116250: Entrez → HGNC symbol
symbols_116250 <- mapIds(
  org.Hs.eg.db,
  keys      = rownames(expr_gse116250),
  column    = "SYMBOL",
  keytype   = "ENTREZID",
  multiVals = "first"
)

# Retain only genes with a unique, non-NA symbol
valid_116250   <- !is.na(symbols_116250) & !duplicated(symbols_116250)
expr_gse116250 <- expr_gse116250[valid_116250, ]
rownames(expr_gse116250) <- symbols_116250[valid_116250]

cat("GSE116250 genes after ID mapping:", nrow(expr_gse116250), "\n")
```

```{r harmonise_ids_gse141910}
# GSE141910: Ensembl → HGNC symbol
symbols_141910 <- mapIds(
  org.Hs.eg.db,
  keys      = rownames(expr_gse141910),
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)

valid_141910   <- !is.na(symbols_141910) & !duplicated(symbols_141910)
expr_gse141910 <- expr_gse141910[valid_141910, ]
rownames(expr_gse141910) <- symbols_141910[valid_141910]

cat("GSE141910 genes after ID mapping:", nrow(expr_gse141910), "\n")
```

## Step 8b: Sample Clustering Per Dataset (Visual Inspection Only)

These dendrograms let you visually assess each dataset before merging.
**Nothing is removed here.** They are provided for reference only.

```{r outlier_dendrogram_gse116250, fig.width=14, fig.height=5}
sampleTree_116250 <- hclust(dist(t(expr_gse116250)), method = "average")

par(cex = 0.6, mar = c(0, 4, 2, 0))
plot(
  sampleTree_116250,
  main = "GSE116250 — sample clustering (visual inspection only)",
  sub  = "", xlab = "", ylab = "Height"
)
```

```{r outlier_dendrogram_gse141910, fig.width=14, fig.height=5}
sampleTree_141910 <- hclust(dist(t(expr_gse141910)), method = "average")

par(cex = 0.6, mar = c(0, 4, 2, 0))
plot(
  sampleTree_141910,
  main = "GSE141910 — sample clustering (visual inspection only)",
  sub  = "", xlab = "", ylab = "Height"
)
```

## Step 9: Combine Matrices

```{r combine_matrices, message=FALSE}
# Find genes present in both datasets
common_genes <- intersect(rownames(expr_gse116250), rownames(expr_gse141910))
cat("Genes common to both datasets:", length(common_genes), "\n")

# Subset both matrices to common genes in the same row order
expr_116250_common <- expr_gse116250[common_genes, ]
expr_141910_common <- expr_gse141910[common_genes, ]

# Merge into a single expression matrix (genes × all samples)
combined_expr <- cbind(expr_116250_common, expr_141910_common)
cat("Combined matrix dimensions (genes × samples):", dim(combined_expr), "\n")

# Build combined metadata
# age and sex were extracted in Step 3; they are carried through here so that
# Step 13 can use them as additional trait columns in the heatmap
combined_meta <- bind_rows(
  pheno_gse116250 %>% dplyr::select(geo_accession, condition, age, sex) %>% mutate(dataset = "GSE116250"),
  pheno_gse141910 %>% dplyr::select(geo_accession, condition, age, sex) %>% mutate(dataset = "GSE141910")
)
rownames(combined_meta) <- combined_meta$geo_accession

cat("Clinical covariate availability:\n")
cat("  Age — non-missing:", sum(!is.na(combined_meta$age)), "/ total:", nrow(combined_meta), "\n")
cat("  Sex — non-missing:", sum(!is.na(combined_meta$sex)), "/ total:", nrow(combined_meta), "\n")

# Confirm columns of expression matrix match rows of metadata
stopifnot(all(colnames(combined_expr) == rownames(combined_meta)))
cat("Expression matrix and metadata alignment confirmed.\n")
```

## Step 10: Batch Correction with ComBat

```{r batch_correction, message=FALSE}
# Batch vector: 1 = GSE116250, 2 = GSE141910
batch <- ifelse(combined_meta$dataset == "GSE116250", 1, 2)
cat("Samples per batch:", table(batch), "\n")

# Model matrix preserving the biological condition
mod <- model.matrix(~ factor(combined_meta$condition))

# ComBat expects genes × samples (combined_expr is already in that orientation)
combined_expr_corrected <- ComBat(
  dat   = combined_expr,
  batch = batch,
  mod   = mod
)
saveRDS(combined_expr_corrected, file = "batch_effect_corrected_matrix.rds")
cat("Batch-corrected matrix dimensions (genes × samples):",
    dim(combined_expr_corrected), "\n")

# Transpose to samples × genes for WGCNA
wgcna_input <- t(combined_expr_corrected)
cat("WGCNA input dimensions (samples × genes):", dim(wgcna_input), "\n")
```

## Step 11: Soft-Thresholding Power Selection

```{r soft_threshold, fig.width=10, fig.height=5}
powers <- c(1:10, seq(12, 30, by = 2))

sft <- pickSoftThreshold(
  wgcna_input,
  powerVector = powers,
  networkType = "signed",
  verbose     = 5
)

par(mfrow = c(1, 2))

plot(
  sft$fitIndices[, 1],
  -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
  xlab = "Soft Threshold (power)",
  ylab = "Scale Free Topology Model Fit (signed R²)",
  type = "n",
  main = "Scale independence"
)
text(
  sft$fitIndices[, 1],
  -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
  labels = powers, cex = 0.9, col = "red"
)
abline(h = 0.85, col = "red", lty = 2)

plot(
  sft$fitIndices[, 1],
  sft$fitIndices[, 5],
  xlab = "Soft Threshold (power)",
  ylab = "Mean Connectivity",
  type = "n",
  main = "Mean connectivity"
)
text(
  sft$fitIndices[, 1],
  sft$fitIndices[, 5],
  labels = powers, cex = 0.9, col = "red"
)
```

```{r soft_threshold_result}
# Show R² and mean connectivity for all tested powers to inform the choice
sft$fitIndices %>%
  mutate(signed_r2 = -sign(slope) * SFT.R.sq) %>%
  dplyr::select(Power, signed_r2, mean.k.) %>%
  print()

# Choose the lowest power where signed R² ≥ 0.85
candidates <- sft$fitIndices %>%
  mutate(signed_r2 = -sign(slope) * SFT.R.sq) %>%
  filter(signed_r2 >= 0.85)

if (nrow(candidates) > 0) {
  cat("Lowest power reaching R² ≥ 0.85:", candidates %>% slice(1) %>% pull(Power), "\n")
} else {
  cat("No power reached R² ≥ 0.85 — selecting manually.\n")
}

# Power 12 is selected by convention for signed networks when no single power
# clearly reaches R² ≥ 0.85, following the WGCNA FAQ recommendation for
# RNA-seq data (Langfelder & Horvath suggest power 12 as a default for
# signed networks when the scale-free criterion is not cleanly met)
soft_power <- 12
cat("Selected soft-thresholding power:", soft_power, "\n")

# Confirm R² and mean connectivity at chosen power
idx       <- which(sft$fitIndices$Power == soft_power)
signed_r2 <- -sign(sft$fitIndices$slope[idx]) * sft$fitIndices$SFT.R.sq[idx]
cat("Scale-free R² at power", soft_power, ":", round(signed_r2, 3), "\n")
cat("Mean connectivity at power", soft_power, ":", round(sft$fitIndices[idx, 5], 1), "\n")
```

## Step 12: Network Construction and Module Detection

`blockwiseModules()` performs the core WGCNA computation in a single call:

1. Raises the correlation matrix to `soft_power` to approximate a scale-free topology
2. Computes the **Topological Overlap Matrix (TOM)** — a measure of how similarly connected two genes are across the whole network (more robust than raw correlation)
3. Clusters genes hierarchically on 1 − TOM distance
4. Cuts the dendrogram into **modules** (branches) using dynamic tree cutting
5. Merges modules whose eigengenes are highly similar (`mergeCutHeight = 0.25` means modules sharing >75% eigengene correlation are collapsed)

Genes that do not cluster into any module are assigned to the **grey module (label 0)** and excluded from downstream analysis.

```{r build_network, message=FALSE, warning=FALSE}
# WGCNA's cor() is used internally; explicitly prefer it over base R's cor()
# to avoid namespace conflicts introduced by loading other packages
conflicts_prefer(WGCNA::cor)

bwnet <- blockwiseModules(
  wgcna_input,
  power             = soft_power,  # soft-thresholding power chosen in Step 11

  # signed network: distinguishes positive (co-expressed) from negative
  # (anti-correlated) gene relationships : appropriate for biological data
  # where direction of co-expression is meaningful
  networkType       = "signed",
  TOMType           = "signed",

  # Module detection parameters
  minModuleSize     = 30,    # modules must contain at least 30 genes
  mergeCutHeight    = 0.25,  # merge modules whose eigengenes correlate > 0.75

  # Process all genes in a single block (avoids artefacts from block-splitting)
  # Requires sufficient RAM; reduce if memory errors occur
  maxBlockSize      = ncol(wgcna_input),

  numericLabels     = TRUE,  # modules labelled 0, 1, 2... (0 = grey/unassigned)
  saveTOMs          = FALSE,
  verbose           = 3
)

# Convert numeric module labels to colour names (WGCNA convention)
# e.g. 0 = "grey", 1 = "turquoise", 2 = "blue", etc.
module_labels  <- bwnet$colors
module_colours <- labels2colors(module_labels)

cat("Number of modules detected (excluding grey):",
    length(unique(module_labels[module_labels != 0])), "\n")
cat("Module sizes (number of genes per module):\n")
print(table(module_colours))
```

```{r plot_dendrogram, fig.width=14, fig.height=6}
# Gene dendrogram: branches represent groups of co-expressed genes.
# The colour bar beneath shows final module assignments after merging.
# Genes in grey are unassigned. Each distinct colour is one module.
plotDendroAndColors(
  bwnet$dendrograms[[1]],
  module_colours[bwnet$blockGenes[[1]]],
  "Module colours",
  dendroLabels = FALSE,
  hang         = 0.03,
  addGuide     = TRUE,
  guideHang    = 0.05,
  main         = "Gene dendrogram and module assignment"
)
```

## Step 13: Module–Trait Correlation

Each module is summarised by its **module eigengene (ME)** — the first principal
component of the expression matrix for all genes in that module. The ME captures the
dominant expression pattern shared across the module's genes and acts as a
representative expression profile for the whole module.

We correlate each ME with four traits:

- **DCM:** the primary biological question — which modules are upregulated or
  downregulated in DCM relative to non-failing hearts?
- **Age:** continuous clinical covariate; identifies modules tracking patient age
  independently of disease status.
- **Sex:** binary clinical covariate (Male = 1, Female = 0); identifies sex-biased
  expression modules.
- **Dataset:** a quality control check — after ComBat batch correction, no module
  should remain strongly correlated with study origin (GSE116250 vs GSE141910).
  A module still tracking dataset after correction likely reflects residual technical
  variation rather than biology.

Age and sex may have missing values for some samples; `use = "pairwise.complete.obs"`
ensures those samples are excluded per trait rather than dropping them globally.

Each cell in the heatmap shows the Pearson **r** (correlation strength, −1 to 1) and
its raw **p-value**. Red = positive correlation; blue = negative correlation.

```{r module_trait_correlation, fig.width=12, fig.height=12}
# Extract module eigengenes computed during blockwiseModules()
# MEs is a samples × modules matrix; each column is one module's eigengene
MEs <- bwnet$MEs

# blockwiseModules returns numeric labels (ME0, ME1, ...); convert to colour names
# so the heatmap rows are labelled e.g. MEturquoise, MEblue, etc.
ME_colours    <- paste0("ME", labels2colors(as.numeric(gsub("ME", "", colnames(MEs)))))
colnames(MEs) <- ME_colours

# Remove MEgrey, the grey module contains unassigned genes and has no
# coherent biological meaning; correlating it with a trait is uninformative
MEs <- MEs[, colnames(MEs) != "MEgrey"]

# Trait data frame: one row per sample, one column per trait
# DCM     = 1 (disease), 0 (control) — primary biological variable
# Age     = continuous (years); may be NA for some samples
# Sex     = 1 (Male), 0 (Female), NA (unknown) — binary clinical covariate
# Dataset = 1 (GSE116250), 0 (GSE141910) — batch sanity check only
trait_df <- data.frame(
  DCM     = ifelse(combined_meta$condition == "DCM", 1, 0),
  Age     = combined_meta$age,
  Sex     = combined_meta$sex,
  Dataset = ifelse(combined_meta$dataset == "GSE116250", 1, 0)
)
rownames(trait_df) <- rownames(combined_meta)
nSamples <- nrow(wgcna_input)

# Pearson correlation between each eigengene and each trait.
# pairwise.complete.obs: for Age and Sex, only samples with non-NA values are
# used when computing those correlations, other samples are not dropped globally.
# corPvalueStudent() converts r to a t-statistic p-value using n − 2 df;
# for Age/Sex the effective n may be smaller due to missing values.
module_trait_cor <- cor(MEs, trait_df, use = "pairwise.complete.obs")
module_trait_p   <- corPvalueStudent(module_trait_cor, nSamples)

# Build the cell text: r on top line, p-value in parentheses below
text_matrix <- paste0(
  round(module_trait_cor, 2), "\n(",
  signif(module_trait_p, 1), ")"
)
dim(text_matrix) <- dim(module_trait_cor)

# Heatmap: each row is a module, columns are DCM and Dataset
# Red  = positive correlation with trait (higher expression in DCM / in GSE116250)
# Blue = negative correlation (lower expression in DCM / in GSE141910)
# White = no correlation
labeledHeatmap(
  Matrix        = module_trait_cor,
  xLabels       = colnames(trait_df),
  yLabels       = colnames(MEs),
  ySymbols      = colnames(MEs),
  colorLabels   = FALSE,
  colors        = blueWhiteRed(50),
  textMatrix    = text_matrix,
  setStdMargins = FALSE,
  cex.text      = 0.6,
  zlim          = c(-1, 1),
  main          = "Module–trait correlations (Pearson r, p-value)"
)
```

```{r significant_modules}
# Filter to modules significantly associated with DCM after BH multiple-testing
# correction. Age, Sex, and Dataset columns are included for contextual interpretation:
#   r_age / r_sex  — check whether a module is driven by a confounding clinical variable
#   r_dataset      — batch sanity check; should be near 0 after ComBat
sig_modules <- data.frame(
  module    = colnames(MEs),
  r_DCM     = round(module_trait_cor[, "DCM"],     3),
  p_value   = signif(module_trait_p[,  "DCM"],     3),
  p_adj     = signif(p.adjust(module_trait_p[, "DCM"], method = "BH"), 3),
  r_age     = round(module_trait_cor[, "Age"],     3),
  r_sex     = round(module_trait_cor[, "Sex"],     3),
  r_dataset = round(module_trait_cor[, "Dataset"], 3)
) %>%
  arrange(p_adj) %>%
  filter(p_adj < 0.05)

cat("Modules significantly correlated with DCM (FDR < 0.05):\n")
print(sig_modules)

# Sanity check: a DCM-significant module with |r_dataset| > 0.3 may be driven
# by study origin rather than true disease biology — interpret with caution
confounded <- sig_modules %>% filter(abs(r_dataset) > 0.3)
if (nrow(confounded) > 0) {
  cat("\nWARNING — these modules may retain batch signal (|r_dataset| > 0.3):\n")
  print(confounded$module)
} else {
  cat("\nNo DCM-significant modules show strong dataset correlation. ComBat successful.\n")
}
```

## Step 14: Hub Gene Identification

**Module membership (kME)** is the Pearson correlation between a gene's expression
profile and the module eigengene. A gene with kME close to 1 is tightly co-expressed
with the rest of the module; a gene with low kME sits on the module periphery.

**Hub genes** are defined as genes with the highest kME — they are the most central
nodes in the co-expression network and represent the strongest candidate regulators or
biomarkers for the biological process captured by each module.

```{r hub_genes}
# signedKME() computes the correlation between every gene and every module eigengene.
# The result is a genes × modules matrix; each column is named kME<colour>.
# Positive kME = gene expression moves with the eigengene (co-expressed with module).
# Negative kME = gene expression moves against the eigengene (anti-correlated).
kME <- signedKME(wgcna_input, MEs, outputColumnName = "kME")

sig_module_names <- sig_modules$module

# For each DCM-significant module, rank all module genes by kME and keep top 20
hub_genes <- lapply(sig_module_names, function(mod) {
  colour  <- gsub("ME", "", mod)
  kme_col <- paste0("kME", colour)  # e.g. kMEturquoise

  data.frame(
    gene   = rownames(kME),
    module = mod,
    kME    = kME[[kme_col]]
  ) %>%
    arrange(desc(kME)) %>%
    slice(1:20)
})

hub_genes_df <- bind_rows(hub_genes)

cat("Top 20 hub genes per significant module (ranked by kME):\n")
print(hub_genes_df)
```

```{r plot_hub_kme, fig.width=15, fig.height=10}
# Restrict plot to the three modules of interest
top3_modules  <- c("MEyellow", "MEred", "MEpurple")
plot_hub_df   <- hub_genes_df %>% dplyr::filter(module %in% top3_modules)

# Dot plot showing kME of the top 20 hub genes per module.
# Genes are sorted by kME (highest at top). Each facet is one module.
# Dot colours match the WGCNA module colour convention.
ggplot(plot_hub_df, aes(x = reorder(gene, kME), y = kME, colour = module)) +
  geom_point(size = 2) +
  facet_wrap(~ module, scales = "free_y", ncol = 3) +
  coord_flip() +
  scale_colour_manual(values = c(MEyellow = "yellow3", MEred = "red", MEpurple = "purple")) +
  labs(
    x     = "Gene",
    y     = "Module membership (kME)",
    title = "Top 20 hub genes per DCM-associated module"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

## Step 15: Gene Significance vs Module Membership

This plot validates the biological relevance of each significant module by comparing
two independent measures for every gene:

- **Gene Significance (GS):** the absolute Pearson correlation between a gene's
  expression and the DCM trait (1 = disease, 0 = control). A high GS means the gene
  is individually dysregulated in DCM, regardless of which module it belongs to.

- **Module Membership (MM / kME):** how tightly a gene co-expresses with its module
  eigengene (from Step 14). A high MM means the gene is a central hub of the module.

**Interpretation:** If a module genuinely captures DCM biology, its hub genes (high MM)
should also be the most disease-associated genes (high GS). A strong positive
correlation between GS and MM within a module (r close to 1) confirms this. A flat or
weak relationship suggests the module has internal co-expression structure but is not
directly relevant to the DCM phenotype.

```{r gs_mm, fig.width=14, fig.height=5}
# Compute Gene Significance: correlate every gene with the binary DCM trait.
# Result is a genes × 1 matrix; convert to data frame for easier handling.
GS_DCM <- cor(wgcna_input, trait_df$DCM, use = "p")
GS_DCM <- as.data.frame(GS_DCM)
colnames(GS_DCM) <- "GS"

# One scatter plot per significant module
gs_mm_plots <- lapply(sig_module_names, function(mod) {
  colour  <- gsub("ME", "", mod)
  kme_col <- paste0("kME", colour)

# Restrict to genes actually assigned to this module
  in_module <- module_colours == colour

  df <- data.frame(
    MM = kME[[kme_col]][in_module],  # module membership for genes in this module
    GS = abs(GS_DCM$GS[in_module])  # absolute GS (direction already captured by r_DCM)
  )

  # Within-module GS–MM correlation: high r validates biological relevance
  r_val <- round(cor(df$MM, df$GS, use = "p"), 2)

  ggplot(df, aes(x = MM, y = GS)) +
    geom_point(alpha = 0.4, size = 0.8, colour = colour) +
    geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
    labs(
      x       = paste0("Module membership (kME, ", colour, ")"),
      y       = "Gene significance (|GS| for DCM)",
      title   = paste0(mod, "  r = ", r_val),
      caption = paste0("n = ", sum(in_module), " genes in module")
    ) +
    theme_bw()
})

# Display all module plots side by side (max 3 per row)
n_cols <- min(3, length(gs_mm_plots))
do.call(gridExtra::grid.arrange, c(gs_mm_plots, ncol = n_cols))
```

```{r gs_mm_summary}
# Summary table: GS–MM correlation coefficient and significance per module.
# High r_GS_MM = hub genes are the most disease-relevant genes = biologically valid.
# Low r_GS_MM  = module structure is real but not driven by DCM phenotype.
gs_mm_stats <- lapply(sig_module_names, function(mod) {
  colour    <- gsub("ME", "", mod)
  kme_col   <- paste0("kME", colour)
  in_module <- module_colours == colour

  mm <- kME[[kme_col]][in_module]
  gs <- abs(GS_DCM$GS[in_module])

  data.frame(
    module   = mod,
    r_GS_MM  = round(cor(mm, gs, use = "p"), 3),
    p_value  = signif(cor.test(mm, gs)$p.value, 3),
    n_genes  = sum(in_module)
  )
})

gs_mm_stats_df <- bind_rows(gs_mm_stats)
cat("GS–MM correlation per significant module:\n")
print(gs_mm_stats_df)
```

#Summary: 

Dilated cardiomyopathy is the most common form of non-ischemic cardiomyopathy 
and the leading indication for heart transplanation worldwide, it typically carries a poor prognosis, with a 50% 5-year mortality without transplant. 
Over 60 genes have been identified to be causative of Dilated Cardiomyopathy but a large proportion of cases remain unexplained that can be improved with processes like WGCNA. 
With this meta analysis, I have combined 2 studies together to identify co-expression modules containing hub genes positively and negatively correlated with the onset of dilated cardiomyopathy. 
Identifying relevant hub genes is essential towards finding novel therapies for Dilated Cardiomyopathy by precisely targeting and interfering with specific portions of the relevant pathways. 
As a preliminary example, inputting the top 20 hub genes within the MEyellow module into EnrichR, suggests that they are involved in Reactome pathways, suggesting coordinated involvement in extracellular matrix remodeling. 
Differential expression analyses can be utilized downstream to compare the differences in expression of a particular gene of a patient to a healthy person. Such analysis can be done for all genes, and upon comparison against the gene modules we can better understand how the genes may relate to each other to fully elucidate the biological roles and pathway interactions of these candidates to determine significance to disease progression. 