---
title: "Cardiomyopathy WGCNA"
author: "Brandon Teo"
date: "2026-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Background

Dilated cardiomyopathy (DCM) is the most common non-ischaemic cardiomyopathy and a leading cause of heart failure and transplantation. Disease progression is driven in part by maladaptive extracellular matrix (ECM) remodelling and cardiac fibrosis, yet the co-ordinated transcriptional programmes underlying these processes remain poorly characterised.
Standard differential expression analysis identifies individual genes that differ between conditions but treats each gene independently, offering limited insight into the higher-order regulatory networks that govern complex phenotypes like fibrosis. Weighted Gene Co-expression Network Analysis (WGCNA) addresses this limitation by grouping genes into modules of co-expressed transcripts and summarising each module with a single eigengene. Module eigengenes can then be correlated with clinical traits, such as disease status, to identify gene programmes collectively associated with DCM. Hub genes within trait-associated modules, defined by high intramodular connectivity, represent candidate regulators of disease-relevant processes.
Thus, we apply WGCNA to bulk RNA-seq data from two independent studies of human left ventricular myocardium (GSE116250 and GSE141910), focusing exclusively on DCM patients and non-failing controls. We construct a co-expression network, identify trait-associated modules, and characterise them through pathway enrichment analysis with particular attention to ECM organisation and TGF-β signalling. This network-based approach complements conventional differential expression by providing a systems-level view of the transcriptomic landscape of DCM-associated cardiac fibrosis.

Research question: What co-expressed gene modules are robustly associated with dilated cardiomyopathy across independent bulk RNA-seq cohorts, and do module-level transcriptional programs implicate fibrotic, inflammatory, or metabolic dysregulation as primary drivers of disease?

```{r load libraries}
library(WGCNA)
library(DESeq2)
library(GEOquery)
library(tidyverse)
library(CorLevelPlot)
library(gridExtra)
```

```{r load datasets}
setwd("~/Desktop/University/NUS research task/Cardiomyopathy WGCNA/")
gse116250 <- getGEO(filename = "GSE116250_series_matrix.txt")
gse141910 <- getGEO(filename = "GSE141910_series_matrix.txt.gz")
getwd()
```

```{r get metadata}
gse116250_phenoData <- pData(phenoData(gse116250))
gse141910_phenoData <- pData(phenoData(gse141910))

# Show all column names so we can identify the disease column
cat("=== GSE116250 columns ===\n"); print(colnames(gse116250_phenoData))
cat("=== GSE141910 columns ===\n"); print(colnames(gse141910_phenoData))

# Show unique values in any characteristics columns
gse116250_char_cols <- grep("characteristics", colnames(gse116250_phenoData), value = TRUE)
for (col in gse116250_char_cols) { cat("\nGSE116250 -", col, ":\n"); print(table(gse116250_phenoData[[col]])) }

gse141910_char_cols <- grep("characteristics", colnames(gse141910_phenoData), value = TRUE)
for (col in gse141910_char_cols) { cat("\nGSE141910 -", col, ":\n"); print(table(gse141910_phenoData[[col]])) }
```

```{r filter samples}
# Find the characteristics column that contains disease labels (not tissue/age/sex)
find_disease_col <- function(phenoData) {
  char_cols <- grep("characteristics", colnames(phenoData), value = TRUE)
  char_cols[sapply(char_cols, function(col) any(grepl("^disease:", phenoData[[col]])))][[1]]
}

# Filter GSE116250: keep only non-failing and dilated cardiomyopathy samples
gse116250_disease_col <- find_disease_col(gse116250_phenoData)
gse116250_keep <- grepl("non.fail|dilated", gse116250_phenoData[[gse116250_disease_col]], ignore.case = TRUE)
gse116250_phenoData <- gse116250_phenoData[gse116250_keep, ]
cat("GSE116250 samples retained:", nrow(gse116250_phenoData), "\n")
table(gse116250_phenoData[[gse116250_disease_col]])

# Filter GSE141910: keep only non-failing and dilated cardiomyopathy samples
gse141910_disease_col <- find_disease_col(gse141910_phenoData)
gse141910_keep <- grepl("non.fail|dilated", gse141910_phenoData[[gse141910_disease_col]], ignore.case = TRUE)
gse141910_phenoData <- gse141910_phenoData[gse141910_keep, ]
cat("GSE141910 samples retained:", nrow(gse141910_phenoData), "\n")
table(gse141910_phenoData[[gse141910_disease_col]])
```

```{r load expression data}
# --- GSE116250: download RPKM supplementary file ---
if (!file.exists("GSE116250_rpkm.txt.gz")) {
  getGEOSuppFiles("GSE116250", makeDirectory = FALSE, baseDir = ".")
}
rpkm_116250 <- read.table("GSE116250_rpkm.txt.gz", header = TRUE, row.names = 1, sep = "\t")

# --- GSE141910: download per-sample count CSVs ---
if (!file.exists("GSE141910_RAW.tar")) {
  getGEOSuppFiles("GSE141910", makeDirectory = FALSE, baseDir = ".")
}
if (!dir.exists("GSE141910_counts")) {
  dir.create("GSE141910_counts")
  untar("GSE141910_RAW.tar", exdir = "GSE141910_counts")
}
csv_files <- list.files("GSE141910_counts", pattern = "\\.csv\\.gz$", full.names = TRUE)
counts_list <- lapply(csv_files, function(f) {
  df <- read.csv(f, header = FALSE, row.names = 1, col.names = c("gene", "count"))
  setNames(df$count, rownames(df))
})
gsm_ids <- sub(".*?(GSM\\d+).*", "\\1", basename(csv_files))
count_matrix_141910 <- do.call(cbind, counts_list)
colnames(count_matrix_141910) <- gsm_ids
```

```{r normalize expression data}
# --- GSE116250: log2(RPKM + 1), filter to retained samples ---
kept_samples_116250 <- rownames(gse116250_phenoData)
rpkm_116250 <- rpkm_116250[, colnames(rpkm_116250) %in% kept_samples_116250, drop = FALSE]

# Filter lowly expressed genes: keep genes with RPKM >= 1 in at least 50% of samples
keep_116250 <- rowSums(rpkm_116250 >= 1) >= (ncol(rpkm_116250) * 0.5)
rpkm_116250 <- rpkm_116250[keep_116250, ]
expr_116250 <- log2(rpkm_116250 + 1)
cat("GSE116250: retained", nrow(expr_116250), "genes across", ncol(expr_116250), "samples\n")

# --- GSE141910: DESeq2 VST on raw counts, filter to retained samples ---
kept_samples_141910 <- rownames(gse141910_phenoData)
count_matrix_141910 <- count_matrix_141910[, colnames(count_matrix_141910) %in% kept_samples_141910, drop = FALSE]
count_matrix_141910 <- count_matrix_141910[rowSums(count_matrix_141910) > 0, ]

col_data_141910 <- data.frame(
  row.names = colnames(count_matrix_141910),
  disease = gse141910_phenoData[colnames(count_matrix_141910), gse141910_disease_col]
)
dds_141910 <- DESeqDataSetFromMatrix(
  countData = count_matrix_141910,
  colData   = col_data_141910,
  design    = ~1
)
keep_141910 <- rowSums(counts(dds_141910) >= 10) >= (ncol(dds_141910) * 0.5)
dds_141910 <- dds_141910[keep_141910, ]
vst_141910 <- vst(dds_141910, blind = TRUE)
expr_141910 <- assay(vst_141910)
cat("GSE141910: retained", nrow(expr_141910), "genes across", ncol(expr_141910), "samples\n")
```

```{r outlier detection}
# Transpose to samples x genes for WGCNA; detect outliers via hierarchical clustering
wgcna_116250 <- t(expr_116250)
wgcna_141910 <- t(expr_141910)

par(mfrow = c(1, 2))
plot(hclust(dist(wgcna_116250), method = "average"), main = "GSE116250 sample clustering",
     xlab = "", sub = "", cex = 0.6)
plot(hclust(dist(wgcna_141910), method = "average"), main = "GSE141910 sample clustering",
     xlab = "", sub = "", cex = 0.6)
par(mfrow = c(1, 1))
# Inspect dendrograms and remove obvious outlier samples before network construction
```

```{r goodSamplesGenes check}
# Check for genes/samples with excessive missing values or zero variance
gsg_116250 <- goodSamplesGenes(wgcna_116250, verbose = 3)
if (!gsg_116250$allOK) {
  wgcna_116250 <- wgcna_116250[gsg_116250$goodSamples, gsg_116250$goodGenes]
  cat("GSE116250: removed", sum(!gsg_116250$goodSamples), "samples and",
      sum(!gsg_116250$goodGenes), "genes\n")
}
cat("GSE116250 final dimensions:", nrow(wgcna_116250), "samples x", ncol(wgcna_116250), "genes\n")

gsg_141910 <- goodSamplesGenes(wgcna_141910, verbose = 3)
if (!gsg_141910$allOK) {
  wgcna_141910 <- wgcna_141910[gsg_141910$goodSamples, gsg_141910$goodGenes]
  cat("GSE141910: removed", sum(!gsg_141910$goodSamples), "samples and",
      sum(!gsg_141910$goodGenes), "genes\n")
}
cat("GSE141910 final dimensions:", nrow(wgcna_141910), "samples x", ncol(wgcna_141910), "genes\n")
```

```{r soft threshold selection, fig.width=12, fig.height=5}
# Test powers 1-20 and pick the lowest power giving scale-free R^2 >= 0.85
powers <- 1:20

sft_116250 <- pickSoftThreshold(wgcna_116250, powerVector = powers,
                                networkType = "signed", verbose = 2)
sft_141910 <- pickSoftThreshold(wgcna_141910, powerVector = powers,
                                networkType = "signed", verbose = 2)

par(mfrow = c(2, 2))

# GSE116250
plot(sft_116250$fitIndices[, 1], -sign(sft_116250$fitIndices[, 3]) * sft_116250$fitIndices[, 2],
     xlab = "Soft Threshold (power)", ylab = "Scale Free Topology R^2",
     main = "GSE116250: Scale-free fit", type = "n")
text(sft_116250$fitIndices[, 1], -sign(sft_116250$fitIndices[, 3]) * sft_116250$fitIndices[, 2],
     labels = powers, col = "red")
abline(h = 0.85, col = "red", lty = 2)

plot(sft_116250$fitIndices[, 1], sft_116250$fitIndices[, 5],
     xlab = "Soft Threshold (power)", ylab = "Mean Connectivity",
     main = "GSE116250: Mean connectivity", type = "n")
text(sft_116250$fitIndices[, 1], sft_116250$fitIndices[, 5], labels = powers, col = "red")

# GSE141910
plot(sft_141910$fitIndices[, 1], -sign(sft_141910$fitIndices[, 3]) * sft_141910$fitIndices[, 2],
     xlab = "Soft Threshold (power)", ylab = "Scale Free Topology R^2",
     main = "GSE141910: Scale-free fit", type = "n")
text(sft_141910$fitIndices[, 1], -sign(sft_141910$fitIndices[, 3]) * sft_141910$fitIndices[, 2],
     labels = powers, col = "red")
abline(h = 0.85, col = "red", lty = 2)

plot(sft_141910$fitIndices[, 1], sft_141910$fitIndices[, 5],
     xlab = "Soft Threshold (power)", ylab = "Mean Connectivity",
     main = "GSE141910: Mean connectivity", type = "n")
text(sft_141910$fitIndices[, 1], sft_141910$fitIndices[, 5], labels = powers, col = "red")

par(mfrow = c(1, 1))

# Print recommended powers
cat("GSE116250 recommended power:", sft_116250$powerEstimate, "\n")
cat("GSE141910 recommended power:", sft_141910$powerEstimate, "\n")
```
