---
title: "Cardiomyopathy WGCNA"
author: "Brandon Teo"
date: "2026-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Background

Dilated cardiomyopathy (DCM) is the most common non-ischaemic cardiomyopathy and a leading cause of heart failure and transplantation. Disease progression is driven in part by maladaptive extracellular matrix (ECM) remodelling and cardiac fibrosis, yet the co-ordinated transcriptional programmes underlying these processes remain poorly characterised.
Standard differential expression analysis identifies individual genes that differ between conditions but treats each gene independently, offering limited insight into the higher-order regulatory networks that govern complex phenotypes like fibrosis. Weighted Gene Co-expression Network Analysis (WGCNA) addresses this limitation by grouping genes into modules of co-expressed transcripts and summarising each module with a single eigengene. Module eigengenes can then be correlated with clinical traits, such as disease status, to identify gene programmes collectively associated with DCM. Hub genes within trait-associated modules, defined by high intramodular connectivity, represent candidate regulators of disease-relevant processes.
Thus, we apply WGCNA to bulk RNA-seq data from two independent studies of human left ventricular myocardium (GSE116250 and GSE141910), focusing exclusively on DCM patients and non-failing controls. We construct a co-expression network, identify trait-associated modules, and characterise them through pathway enrichment analysis with particular attention to ECM organisation and TGF-β signalling. This network-based approach complements conventional differential expression by providing a systems-level view of the transcriptomic landscape of DCM-associated cardiac fibrosis.

Research question: What co-expressed gene modules are robustly associated with dilated cardiomyopathy across independent bulk RNA-seq cohorts, and do module-level transcriptional programs implicate fibrotic, inflammatory, or metabolic dysregulation as primary drivers of disease?

---

## Step 1: Load Libraries

```{r install_bioc_packages, message=FALSE, warning=FALSE}
# Install Bioconductor packages if not already present
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

bioc_pkgs <- c("WGCNA", "DESeq2", "GEOquery", "AnnotationDbi", "org.Hs.eg.db", "sva")
for (pkg in bioc_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE))
    BiocManager::install(pkg)
}
```

```{r load_libraries, message=FALSE, warning=FALSE}
library(WGCNA)
library(DESeq2)
library(GEOquery)
library(tidyverse)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(sva)
library(CorLevelPlot)
library(gridExtra)

# Allow multi-threading for WGCNA
allowWGCNAThreads()
```

---

## Step 2: Load Metadata (Series Matrices)

```{r load_metadata, message=FALSE, warning=FALSE}
setwd("~/Desktop/University/NUS research task/Cardiomyopathy WGCNA/")

# Load GEO series matrix files — these contain sample-level metadata only
gse116250_geo <- getGEO(filename = "GSE116250_series_matrix.txt",    GSEMatrix = TRUE)
gse141910_geo <- getGEO(filename = "GSE141910_series_matrix.txt.gz", GSEMatrix = TRUE)

# Extract phenotypic data tables (rows = samples, columns = metadata fields)
pheno_gse116250 <- pData(phenoData(gse116250_geo))
pheno_gse141910 <- pData(phenoData(gse141910_geo))

cat("GSE116250 total samples:", nrow(pheno_gse116250), "\n")
cat("GSE141910 total samples:", nrow(pheno_gse141910), "\n")
```

---

## Step 3: Filter Metadata to DCM and Non-Failing Samples Only

```{r filter_metadata}
# --- GSE116250 ---
# Condition is stored in the 'disease:ch1' column
# Values: "dilated cardiomyopathy", "ischemic cardiomyopathy", "non-failing"
cat("GSE116250 disease groups:\n")
print(table(pheno_gse116250[["disease:ch1"]]))

pheno_gse116250 <- pheno_gse116250 %>%
  filter(`disease:ch1` %in% c("dilated cardiomyopathy", "non-failing")) %>%
  mutate(condition = ifelse(`disease:ch1` == "non-failing", "NF", "DCM"))

cat("\nGSE116250 after filter:", nrow(pheno_gse116250), "samples\n")
print(table(pheno_gse116250$condition))

# --- GSE141910 ---
# Condition is stored in the 'etiology:ch1' column
# Values: "Dilated cardiomyopathy (DCM)", "Non-Failing Donor",
#         "Hypertrophic cardiomyopathy (HCM)", "Peripartum cardiomyopathy (PPCM)"
cat("\nGSE141910 etiology groups:\n")
print(table(pheno_gse141910[["etiology:ch1"]]))

pheno_gse141910 <- pheno_gse141910 %>%
  filter(`etiology:ch1` %in% c("Dilated cardiomyopathy (DCM)", "Non-Failing Donor")) %>%
  mutate(condition = ifelse(`etiology:ch1` == "Non-Failing Donor", "NF", "DCM"))

cat("\nGSE141910 after filter:", nrow(pheno_gse141910), "samples\n")
print(table(pheno_gse141910$condition))
```

---

## Step 4: Load Count Matrices

### 4a. GSE116250 — Pre-assembled raw integer counts

```{r load_gse116250_counts}
# This file is a tab-separated matrix:
#   rows    = genes (Entrez IDs)
#   columns = samples (GSM accession IDs)
#   values  = raw read counts from STAR/featureCounts
counts_gse116250 <- read.table(
  "GSE116250_raw_counts_GRCh38.p13_NCBI.tsv",
  header      = TRUE,
  sep         = "\t",
  row.names   = 1,
  check.names = FALSE
)

cat("GSE116250 count matrix dimensions:", dim(counts_gse116250), "\n")
cat("Gene ID format (Entrez):", head(rownames(counts_gse116250), 3), "\n")
cat("Sample ID format (GSM):", head(colnames(counts_gse116250), 3), "\n")
cat("Values are integers:", !any(as.matrix(counts_gse116250) != round(as.matrix(counts_gse116250)), na.rm = TRUE), "\n")
```

### 4b. GSE141910 — Assemble from individual per-sample files

**Note:** The GEO-deposited files for GSE141910 contain Voom-transformed log2 values
(confirmed by GEO metadata: *"Raw gene counts were transformed using SVA and the
LIMMA-Voom method"*). Raw STAR counts were not deposited. These pre-normalised values
will be used directly for WGCNA after filtering; DESeq2 will not be applied to this
dataset.

```{r assemble_gse141910_counts, message=FALSE}
folder_path <- "GSE141910_counts"
file_list   <- list.files(path = folder_path, pattern = "\\.gz$", full.names = TRUE)
cat("Number of sample files found:", length(file_list), "\n")

# Function to read one sample file and name its count column with the GSM accession ID
read_and_name <- function(file) {
  temp_data   <- read.csv(gzfile(file), header = TRUE)
  # Extract the GSM accession ID from the filename (everything before the first underscore)
  sample_name <- gsub("(_.*|\\.csv\\.gz|\\.gz)", "", basename(file))
  colnames(temp_data) <- c("Gene", sample_name)
  return(temp_data)
}

# Read all files and merge into a single matrix by the Gene column
all_samples_list <- lapply(file_list, read_and_name)
counts_gse141910 <- Reduce(
  function(x, y) merge(x, y, by = "Gene", all = TRUE),
  all_samples_list
)

# Promote gene IDs to row names
rownames(counts_gse141910) <- counts_gse141910$Gene
counts_gse141910 <- counts_gse141910[, -1]

cat("GSE141910 count matrix dimensions:", dim(counts_gse141910), "\n")
cat("Gene ID format (ENSEMBL):", head(rownames(counts_gse141910), 3), "\n")
cat("Sample ID format (GSM):", head(colnames(counts_gse141910), 3), "\n")
```

---

## Step 5: Align Count Matrices with Filtered Metadata

```{r align_gse116250}
# --- GSE116250 ---
# Metadata rownames and count matrix column names are both GSM accession IDs
cat("GSE116250 — samples in metadata:", nrow(pheno_gse116250), "\n")
cat("GSE116250 — samples in count matrix:", ncol(counts_gse116250), "\n")
cat("GSE116250 — matching samples:",
    sum(colnames(counts_gse116250) %in% rownames(pheno_gse116250)), "\n")

# Subset count matrix to filtered samples only, in metadata row order
counts_gse116250 <- counts_gse116250[, rownames(pheno_gse116250)]

# Sanity check: column order must exactly match metadata row order
stopifnot(all(colnames(counts_gse116250) == rownames(pheno_gse116250)))
cat("GSE116250 alignment check passed.\n")
```

```{r align_gse141910}
# --- GSE141910 ---
cat("GSE141910 — samples in metadata:", nrow(pheno_gse141910), "\n")
cat("GSE141910 — samples in count matrix:", ncol(counts_gse141910), "\n")
cat("GSE141910 — matching samples:",
    sum(colnames(counts_gse141910) %in% rownames(pheno_gse141910)), "\n")

counts_gse141910 <- counts_gse141910[, rownames(pheno_gse141910)]

stopifnot(all(colnames(counts_gse141910) == rownames(pheno_gse141910)))
cat("GSE141910 alignment check passed.\n")
```

---

## Step 6: Normalise with DESeq2 (GSE116250 only)

GSE116250 contains genuine raw integer counts and is processed through the full DESeq2
pipeline. Variance stabilising transformation (VST) produces approximately log2-scale
values suitable for WGCNA.

```{r deseq2_gse116250, message=FALSE, warning=FALSE}
# Build the DESeqDataSet with condition as the design variable
dds116250 <- DESeqDataSetFromMatrix(
  countData = counts_gse116250,
  colData   = pheno_gse116250[, "condition", drop = FALSE],
  design    = ~ condition
)

# Set NF as the reference level
dds116250$condition <- relevel(factor(dds116250$condition), ref = "NF")

# Pre-filter: remove genes with fewer than 10 counts in at least 50% of samples
# This reduces memory and improves statistical power
keep <- rowSums(counts(dds116250) >= 10) >= ceiling(ncol(dds116250) * 0.5)
dds116250 <- dds116250[keep, ]
cat("GSE116250 genes after low-count filter:", nrow(dds116250), "\n")

# Run DESeq2 (estimates size factors, dispersions, and fits the model)
dds116250 <- DESeq(dds116250)

# Apply VST (blind = TRUE: do not use the design for transformation,
# appropriate when combining datasets)
vst116250    <- vst(dds116250, blind = TRUE)
expr_gse116250 <- assay(vst116250)

cat("GSE116250 VST expression matrix dimensions:", dim(expr_gse116250), "\n")
cat("Value range:", round(min(expr_gse116250), 2), "—", round(max(expr_gse116250), 2), "\n")
```

---

## Step 7: Filter GSE141910 Pre-normalised Values

GSE141910 Voom values are on a log2 scale. We apply an expression filter analogous to
the count filter above (retain genes expressed above log2(10+1) ≈ 3.46 in ≥50% of
samples).

```{r filter_gse141910}
keep141910   <- rowSums(counts_gse141910 >= log2(10 + 1)) >= ceiling(ncol(counts_gse141910) * 0.5)
expr_gse141910 <- as.matrix(counts_gse141910[keep141910, ])

cat("GSE141910 genes after expression filter:", nrow(expr_gse141910), "\n")
cat("Value range:", round(min(expr_gse141910), 2), "—", round(max(expr_gse141910), 2), "\n")
```

---

## Step 8: Harmonise Gene IDs to HGNC Gene Symbols

GSE116250 uses Entrez IDs; GSE141910 uses Ensembl IDs. Both are mapped to HGNC symbols
so the two matrices share a common identifier for merging.

```{r harmonise_ids_gse116250}
# GSE116250: Entrez → HGNC symbol
symbols_116250 <- mapIds(
  org.Hs.eg.db,
  keys      = rownames(expr_gse116250),
  column    = "SYMBOL",
  keytype   = "ENTREZID",
  multiVals = "first"
)

# Retain only genes with a unique, non-NA symbol
valid_116250   <- !is.na(symbols_116250) & !duplicated(symbols_116250)
expr_gse116250 <- expr_gse116250[valid_116250, ]
rownames(expr_gse116250) <- symbols_116250[valid_116250]

cat("GSE116250 genes after ID mapping:", nrow(expr_gse116250), "\n")
```

```{r harmonise_ids_gse141910}
# GSE141910: Ensembl → HGNC symbol
symbols_141910 <- mapIds(
  org.Hs.eg.db,
  keys      = rownames(expr_gse141910),
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)

valid_141910   <- !is.na(symbols_141910) & !duplicated(symbols_141910)
expr_gse141910 <- expr_gse141910[valid_141910, ]
rownames(expr_gse141910) <- symbols_141910[valid_141910]

cat("GSE141910 genes after ID mapping:", nrow(expr_gse141910), "\n")
```

---

## Step 9: Combine Matrices and Correct for Batch Effects

```{r combine_and_batch_correct, message=FALSE}
# Find genes present in both datasets
common_genes <- intersect(rownames(expr_gse116250), rownames(expr_gse141910))
cat("Genes common to both datasets:", length(common_genes), "\n")

# Subset both matrices to common genes (same row order)
expr_116250_common <- expr_gse116250[common_genes, ]
expr_141910_common <- expr_gse141910[common_genes, ]

# Merge into a single expression matrix (genes × all samples)
combined_expr <- cbind(expr_116250_common, expr_141910_common)
cat("Combined matrix dimensions:", dim(combined_expr), "\n")

# Build batch vector: 1 = GSE116250, 2 = GSE141910
batch <- c(
  rep(1, ncol(expr_116250_common)),
  rep(2, ncol(expr_141910_common))
)

# Build combined condition vector for the ComBat model matrix
combined_condition <- factor(c(pheno_gse116250$condition, pheno_gse141910$condition))
mod <- model.matrix(~ combined_condition)

# Apply ComBat to remove batch-driven variance while preserving biological signal
combined_expr_corrected <- ComBat(
  dat   = combined_expr,
  batch = batch,
  mod   = mod
)

cat("Batch-corrected matrix dimensions:", dim(combined_expr_corrected), "\n")

# Build combined metadata for downstream WGCNA trait correlation
combined_meta <- bind_rows(
  pheno_gse116250 %>% select(geo_accession, condition) %>% mutate(dataset = "GSE116250"),
  pheno_gse141910 %>% select(geo_accession, condition) %>% mutate(dataset = "GSE141910")
)
rownames(combined_meta) <- combined_meta$geo_accession
```

---

## Step 10: Prepare for WGCNA

WGCNA expects the expression matrix in **samples × genes** orientation.

### 10a. Quality check — remove bad genes and samples

```{r wgcna_qc}
# Transpose: rows = samples, columns = genes
wgcna_input <- t(combined_expr_corrected)

# Identify and remove genes or samples with too many missing values or zero variance
gsg <- goodSamplesGenes(wgcna_input, verbose = 3)

if (!gsg$allOK) {
  cat("Removing", sum(!gsg$goodGenes), "bad genes and",
      sum(!gsg$goodSamples), "bad samples.\n")
  wgcna_input <- wgcna_input[gsg$goodSamples, gsg$goodGenes]
}

cat("Final WGCNA input dimensions (samples × genes):", dim(wgcna_input), "\n")
```

### 10b. Sample outlier detection via hierarchical clustering

```{r sample_outlier_plot, fig.width=14, fig.height=6}
sampleTree <- hclust(dist(wgcna_input), method = "average")

par(cex = 0.6, mar = c(0, 4, 2, 0))
plot(
  sampleTree,
  main  = "Sample clustering to detect outliers",
  sub   = "",
  xlab  = "",
  ylab  = "Height"
)

# Optional: draw a cut line to flag outliers (adjust height as needed after inspection)
# abline(h = 80, col = "red")
```

### 10c. Soft-thresholding power selection

```{r soft_threshold, fig.width=10, fig.height=5}
powers <- c(1:10, seq(12, 20, by = 2))

sft <- pickSoftThreshold(
  wgcna_input,
  powerVector = powers,
  networkType = "signed",
  verbose     = 5
)

# Plot 1: scale-free topology model fit (R²)
# Plot 2: mean network connectivity
par(mfrow = c(1, 2))

plot(
  sft$fitIndices[, 1],
  -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
  xlab = "Soft Threshold (power)",
  ylab = "Scale Free Topology Model Fit (signed R²)",
  type = "n",
  main = "Scale independence"
)
text(
  sft$fitIndices[, 1],
  -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
  labels = powers,
  cex    = 0.9,
  col    = "red"
)
abline(h = 0.85, col = "red", lty = 2)

plot(
  sft$fitIndices[, 1],
  sft$fitIndices[, 5],
  xlab = "Soft Threshold (power)",
  ylab = "Mean Connectivity",
  type = "n",
  main = "Mean connectivity"
)
text(
  sft$fitIndices[, 1],
  sft$fitIndices[, 5],
  labels = powers,
  cex    = 0.9,
  col    = "red"
)
```

```{r soft_threshold_result}
# Choose the lowest power where R² ≥ 0.85
soft_power <- sft$fitIndices %>%
  mutate(signed_r2 = -sign(slope) * SFT.R.sq) %>%
  filter(signed_r2 >= 0.85) %>%
  slice(1) %>%
  pull(Power)

cat("Selected soft-thresholding power:", soft_power, "\n")
```
